---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Overview of HVPSUM

<!-- badges: start -->
<!-- badges: end -->

**HVPSUM** is an R package designed to provide unbiased estimates of heritability and genetic correlation by disentangling horizontal and vertical pleiotropy using summary data. Unlike Linkage Disequilibrium Score Regression (LDSC), the HVPSUM model adjusts for vertical pleiotropy by incorporating causal effect estimate from Mendelian Randomization (MR). This enables a clear distinction between:

- **Shared genetic influences** (horizontal pleiotropy), and
- **Genetic effects mediated through causal pathways** (vertical pleiotropy).

By addressing the limitations of conventional approaches, the HVPSUM package offers a robust framework for analyzing complex genetic relationships.

## Installation

You can install the development version of HVPSUM from [GitHub](https://github.com/) with:

```r
# install.packages("devtools")
devtools::install_github("lamessad/HVPSUM")
```

# Steps for the Horizontal and Vertical Pleiotropy (HVP) Model based on summary data

## 1. Munge the Summary Data (see LDSC for details)

## 2. Run MR

- Estimate the causal effect size (`tau`) of the exposure on the outcome using robust MR methods (e.g., MRLOVA).
- Use the **same summary dataset** to ensure consistency between the LDSC analysis and MR.

## 3. Apply the HVPSUM

This R package performs the following steps:

1. **LDSC Mode**: If `ldsc_env_path`, `ldsc_exe_path`, `ld_path`, and munged `summary data` are provided, the package performs LDSC analysis to estimate heritability and genetic covariance for each block using jackknife resampling. The required inputs are:
   - `ldsc_env_path`: Path to the LDSC environment.
   - `ldsc_exe_path`: Path to the LDSC executable.
   - `ld_path`: Path to the reference LD data.
   - `sumstats_outcome`: Path to the munged summary statistics file for the outcome trait.
   - `sumstats_exposure`: Path to the munged summary statistics file for the exposure trait.

   When these paths are provided, the package will execute LDSC analysis and compute the necessary block estimates from the jackknife resampling.
 
   ```r
   # Example using LDSC Mode
   #library(HVPSUM)
   #hvpsum(ldsc_env_path = "/data/alh-admlda/anaconda3/envs/ldsc/bin/python", 
   #        ldsc_exe_path = "/data/alh-admlda/ldsc/ldsc.py",                   
   #        ld_path = "/data/alh-admlda/ldsc/eur_w_ld_chr/", 
   #        sumstats_outcome = "/data/alh-admlda/DM2.sumstats.gz", 
   #        sumstats_exposure = "/data/alh-admlda/mets.sumstats.gz", 
   #        tau = 0.21)
   ```

2. **Pre-computed Mode**: If pre-computed data are provided for `h2_outcome`, `h2_exposure`, and `gencov`, these are used directly to bypass the LDSC analysis. The required inputs are:

   - `h2_outcome`: Pre-computed heritability estimates for each block for the outcome.
   - `h2_exposure`: Pre-computed heritability estimates for each block for the exposure.
   - `gencov`: Pre-computed genetic covariance estimates for each block.

   The outputs include estimates, standard errors, and p-values for corrected and uncorrected parameters.

   ```{r}
#Example using Pre-computed Mode

   library(HVPSUM)
   data("dat1")
   data("dat2")
   data("dat3")
   head(dat1)
   head(dat2)
   head(dat3)
   tau <- 0.21
   hvpsum(h2_outcome = dat1, h2_exposure = dat2, gencov = dat3, tau =tau)   
   ```

## Contact

Please contact Lamessa Amente (lamessa.amente@mymail.unisa.edu.au) or Hong Lee (hong.lee@unisa.edu.au) if you have any queries.
